---
title: "Using app(), lapp(), and tapp() with SpatRaster in R"
author: "Philippe Massicotte"
date: "2025-06-12"
categories: [R, terra, gis]
editor_options:
  chunk_output_type: console
citation: true
---

```{r}
#| label: renv
#| include: false

renv::use(lockfile = "renv.lock")
```

```{r}
#| label: setup
#| include: false

library(terra)
```

In this short post, I will show how to use the `app()`, `lapp()`, and `tapp()` functions from the `terra` package to apply functions to `SpatRaster` objects. These functions are useful for performing operations on raster data, such as calculating means or applying custom functions across layers.

![Aerial view of layered rock formations showing natural geological strata, representing the concept of working with raster layers in spatial data analysis. Photo by [Eberhard Gross‑Gasteiger](https://www.pexels.com/@eberhardgross/) on Pexels.](https://images.pexels.com/photos/443446/pexels-photo-443446.jpeg){.preview-image}

## Creating a `SpatRaster`

```{r}
#| label: create-spatraster

library(terra)

# Create a SpatRaster with 3 layers, 2x2 pixels
r <- rast(nrows = 2L, ncols = 2L, nlyrs = 3L)

# Fill with numbers 1 to 12
values(r) <- 1L:12L

# Assign names to the layers
names(r) <- c("lyr1", "lyr2", "lyr3")

r

r[]
```

```{r}
#| label: plot-spatraster

plot(r, main = names(r), col = terrain.colors(12L))
```

## Applying functions to `SpatRaster` layers

### Using `app()`

Use `app()` to apply a function across all layers of a `SpatRaster` object. This function operates on each pixel across all layers, allowing you to compute statistics or perform operations on the raster data.

```{r}
#| label: app-mean

# Apply mean across layers for each pixel
app(r, mean)[]
```

### Using `lapp()`

Use `lapp()` to apply a custom function that takes multiple layers as input. This is useful for more complex calculations that involve multiple layers.

```{r}
#| label: lapp-custom

# Apply custom function across layers
lapp(r, function(a, b, c) a + 2L * b + c)[]
```

### Using `tapp()`

Use `tapp()` to apply a function over groups of layers, which is useful for temporal data or when you want to summarize layers by groups.

```{r}
#| label: tapp-groups

# Group layers and apply mean (layers 1 and 2 are grouped, layer 3 separate)
tapp(r, mean, index = c(1L, 1L, 2L))[]
```

### Using `global()`

Use the `global()` function to computes a summary statistic for each layer across all pixels. This is useful for obtaining overall statistics like `mean`, `sum`, or standard deviation for each layer.

```{r}
#| label: global-mean

# Calculate global mean for each layer
global(r, mean)
```

## Functions comparison

::: {.column-page}

| Function&nbsp;&nbsp;&nbsp; | Operates Over       | Purpose                                      | Output Type              | Example Usage                          | When to Use                                                        |
| :------------------------- | ------------------- | -------------------------------------------- | ------------------------ | -------------------------------------- | ------------------------------------------------------------------ |
| `app()`                    | Layers (per pixel)  | Apply a function across layers at each pixel | `SpatRaster` (1+ layers) | `app(r, mean)`<br>`app(r, sum)`        | Simple summaries across layers (`mean`, `sum`, `min`, etc.)        |
| `lapp()`                   | Layers (vectorized) | Pixel-wise operations with vectorized inputs | `SpatRaster`             | `lapp(r, function(a,b,c) a + 2*b - c)` | Custom functions, better performance, multi-input, per-layer logic |
| `tapp()`                   | Time groups         | Aggregate over temporal/index groups         | `SpatRaster`             | `tapp(r, mean, index = c(1,1,2))`      | Temporal summaries like monthly or yearly mean, grouped ops        |
| `global()`                 | Cells (per layer)   | Summary over all pixels in each layer        | `data.frame`             | `global(r, mean)`                      | Whole-layer stats like `mean`, `min`, `sum`, `sd`, etc.            |

:::

<details>
<summary>Session info</summary>

```{r}
#| echo: false
#| label: sessioninfo

devtools::session_info()
```

</details>

<details>
<summary>renv.lock file</summary>

```{.json include="renv.lock"}

```

</details>
