---
title: "Separating/exploding spatial features with R"
author: "Philippe Massicotte"
date: "2023-08-15"
categories: [R, rspatial, Geospatial]
# the preview file needs to be named preview.png
# mogrify -format png preview.jpg
# https://quarto.org/docs/websites/website-tools.html#preview-images
editor_options:
  chunk_output_type: console
citation: true
---

```{r}
#| label: renv
#| include: false
# https://www.joelnitta.com/posts/2024-01-11_using_renv_with_blog/

renv::use(lockfile = "renv.lock")

library(quarto)
```

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(ggpmthemes)
library(sf)
library(rnaturalearthhires)
theme_set(theme_minimal(base_family = "montserrat"))
```

It has been a while since I posted something! But today I wanted to quickly experiment and create a map where I could separate/explode the spatial geometries. Let's first download the polygon data for Canada.

```{r}
#| label: load-data
canada <- rnaturalearth::ne_states(country = "canada", returnclass = "sf") |>
  st_make_valid()
```

As we can see, there are three different regions.

```{r}
#| label: explore-regions
canada |>
  distinct(region)
```

These regions can be visualized using `ggplot2` and `sf`.

```{r}
#| label: plot-original-map
# Get a good CRS for Canada
crsuggest::suggest_crs(canada)

canada <- canada |>
  st_transform(5931)

canada |>
  ggplot(aes(fill = region)) +
  geom_sf()
```

With this following code, we can find the centroid of each region and shift the centers of the new polygons.

```{r}
#| label: explode-geometries
canada_exploded <- canada |>
  group_by(region) |>
  mutate(region_centroid = st_centroid(st_union(geometry))) |>
  ungroup() |>
  mutate(geometry = (geometry - region_centroid) * 0.9 + region_centroid)

st_crs(canada_exploded)
```

As seen above, for a reason I do not know, we have lost the CRS after the operation, so I have to set it back before visualizing it.

```{r}
#| label: plot-exploded-map
canada_exploded <- canada_exploded |>
  st_set_crs(5931)

canada_exploded |>
  ggplot(aes(fill = region)) +
  geom_sf()
```

<details>
  
<summary>Session info</summary>

```{r sessioninfo, echo = FALSE}
## Reproducibility info
options(width = 120)
devtools::session_info()
```

</details>
